\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza

### 2020-02-28 paper/SiM-v1/draft_v16.tex
added h1n1 data

\noindent \textbf{Gaussian-Monotone-PAVA (G-M) ($\hat{X}_{GP}$, $\hat{Y}_{GP}$, $\hat{Z}_{GP}$)}
\begin{align*}
 X_{GP}(t) &= \textnormal{PAVA}(X_G(t)) \\
  Y_{GP}(t) &= N - X_{GP}(t) - Z_{GP}(t) \\
  Z_{GP}(t) &= \textnormal{PAVA}(Z_G(t)) \\
\end{align*}
\textbf{Autoregressive-Monotone-PAVA (AR-M) ($\hat{X}_{ARP}$, $\hat{Y}_{ARP}$, $\hat{Z}_{ARP}$)}
\begin{align*}
  X_{ARP}(t) &= \textnormal{PAVA}(X_{AR}(t)) \\
  Y_{ARP}(t) &= N - X_{ARP}(t) - Z_{ARP}(t) \\
  Z_{ARP}(t) &= \textnormal{PAVA}(Z_{AR}(t)) 
\end{align*}
This gives us a total of four error structures for each set of conditions  (1) Gaussian (G), (2) Gaussian monotone - PAVA (G-M),  (3) autoregressive (AR), and (4) autoregressive monotone - PAVA (AR-M) for a given set of parameters $\left (\beta, \gamma, T, X(0), Y(0), Z(0), \sigma_X, \sigma_Y, N \right )$, the infection rate; recovery rate; total number of time points; initial conditions for number of susceptibles, infectious, and recovered individuals; the standard deviations for both the susceptible and infectious compartments; and the total population, respectively.

\subsection{SIR Data}

We first simulate a ``baseline'' data set from the SIR model under the following conditions:

\noindent \textbf{Baseline Simulated Data}
\begin{center}

	$\beta$ = 0.06, $\gamma$ = 0.03 ($\rr$ = 2)

	Number of time steps $T$: 365

	Starting X size ($X(0)$): 99950

	Starting Y size ($Y(0)$): 50

	Starting Z size ($Z(0)$): 0

	$\sigma_X$: 100

	$\sigma_Y$: 5

	Total population $N$: 100,000

        Error: Gaussian.

      \end{center}
      The baseline data set is generated by adding noise to the susceptible and infectious curves from deterministic SIR paths, which are shown in Figure \ref{fig:baseline-data}.  The recovered compartment is then adjusted to maintain a constant population value.  In this scenario, we see that the infection begins slowly but over the course of a year infects nearly 75\% of the population.
      \begin{figure}
        \centering
        \includegraphics[width=.8\textwidth]{images/baseline_sir.pdf}
        \caption{SIR (XYZ) curves for the deterministic model with $N=100,000$, $(\beta, \gamma) = (0.06, 0.03)$, $(X(0), Y(0), Z(0))= (99950, 50,0)$  The baseline data is generated by using this model and adding noise to the $X$ and $Y$ compartments and adjusting $Z$ at each time step to maintain a constant population.}\label{fig:baseline-data}
        \end{figure}
      This baseline data set is chosen due to its moderate $\rr$ size of 2, which indicates an outbreak of an epidemic similar to ZIK-V, EVD, and H1N1 pandemic influenza (see Table \ref{tab:r0-real-ex}).  The infection rate and recovery rate ($\beta, \gamma$) are relatively small.  The duration of the disease, under the true model, spans nearly a whole year (see Figure \ref{fig:baseline-data}) and so we use a year's worth of daily observations of ($\hat{X},\hat{Y}, \hat{Z}$) values.  Initially, 50 (0.05\%) infectious individuals are introduced to the population of 100,000 individuals (approximately the average population size of a US county in 2015).  The standard deviation of the $\hat{X}$ values is 0.1\% of the magnitude of the initial number of susceptible individuals, and the standard deviation of the $\hat{Y}$ values is 10\% of the magnitude of the initial number of infectious individuals, although in this setting, the number of infectious individuals increases rapidly.  We use Gaussian error as our baseline error because of its well known statistical properties so we can more easily compare methods to one another.  Later on, we explore the autoregressive and PAVA simulated data, which may better simulate data from  a real epidemic.

@@

After generating our baseline data set, we  generate different SIR data sets, changing one of the conditions for simulation each time and using the baseline SIR data as our basis for comparison. The different data sets for a given error structure are shown in Table \ref{tab:simulated-data}.  We repeat  Table \ref{tab:simulated-data}, replacing the ``Error'' feature with  one of the other three error structures for a total of 96 simulated data sets.

@@

\section{Results}\label{sec:results-top}
\subsection{$\rr$ estimation methods}\label{sec:results}
There are numerous aspects to consider when using a method to estimate $\rr$.  Among these include values of both the infection and recovery rate $(\beta, \gamma)$, the number of data points available ($T$), the total size of the population ($N$), the initial percent of susceptible individuals $\left (\frac{X(0)}{N}\right)$, the initial percent of infectious individuals $\left (\frac{Y(0)}{N}\right )$, the magnitude of the variance of both the number of susceptibles and the number of infectious ($\sigma_X^2, \sigma_Y^2$), the error modeling assumptions, data generated non-SIR models, and finally methods of estimating the variance.  For the sake of brevity, we do not provide the full results here, and instead examine the results at a high-level.  However, for those interested readers, we provide results in the form of forest plots and tables of estimates of $\rr$ for each data set in the Supplementary Material.

We implement the \wxxsir methods to estimate $\rr$ using the baseline data set with normal error.  The results are displayed in Figure \ref{fig:baseline-res} and Table \ref{tab:baseline-res}.
\begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/BaseBase.jpg}
  \caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data set $(\beta=.06, \gamma=.03, O=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical, dashed line is the true $\rr$ value.}\label{fig:baseline-res}
  \end{figure}

@@

Overall, in Figure \ref{fig:baseline-res}, we see a wide variation in the estimates of and the standard errors for $\rr$ depending on the method used.  The methods of RE, rRE, LMA, LMAT, IPR, and LL all result in a CI that covers the true value of $\rr=2$.  Of these, LMAT and IPR result in a CI which contains 1, so we would not be able to say whether or the outbreak of a disease was due to random chance or not. On the other hand, many methods appear to underestimate the true $\rr$, and many of them have larger standard errors as well.  We would say that IPR's CI is uninformative. Additionally, MC has a standard error of less than $.0001$, which is very small.  LBE, which assumes that the X and Z compartments are monotonic, overestimates $\rr$

From examining the full results in the Supplementary Material, the results of the baseline data set with normal error seem to be a good summary of the methods as a whole.  In Figure \ref{fig:baseline-res}, we see that the clear best methods to estimate $\rr$ are RE, rRE, and LL.  These three methods consistently outperform the other methods of accurately estimating $\rr$ with a reasonably small CI when varying a single parameter and holding the others as fixed from the generated data sets.

In particular, MC and SB almost never capture the true estimate and typically underestimate the true value of $\rr$.  This leads us to believe that these models are inadequate in capturing the effects of a disease following the SIR model.  Both models rely on parametric assumptions of simple underlying statistical distributions for the course of infection (Binomial and Poisson, respectively) with time-invariant transmission parameters ($\alpha$ and $\lambda$).  These are good examples of why we need to consider other reasons for using a model other than the math resulting in a simple estimate.  That is not to say that variants of these models are unviable for modelling $\rr$, but they do not make for good defaults.  It is also an issue that these two methods consistently underestimate $\rr$, which could possibly lead to an unchecked epidemic.

On the other hand, LBE also tends to have larger CIs and does not often capture the true value of $\rr$ when the the X and Z compartments are not monotonic functions.  This method tends to overestimate the value of $\rr$.

@@

Of the three well-performing methods, RE and rRE seem to be slightly more robust to parameter variation than LL.  This is mostly due to the fact that LL produces such small CIs that even a small discrepancy in estimation of $\rr$ can cause the CI to not capture the true value of $\rr$.

However, we do see instances where LL completely outperforms RE and rRE such as in the case where the initial percent of infectious $\frac{I(0)}{N} \cdot 100\%$ is $< .01$\%. In Figure \ref{fig:ar-small-I}, we examine the data set generated with autoregressive monotonic error, our most stringent data structure, with varying levels of initial percent of infectious individuals.  When $\frac{I(0)}{N} \cdot 100\%$ $< .01$\%, RE and rRE tend to overestimate $\rr$ and have large CIs.  However, LL accurately estimates $\rr$ even when the initial percentage of infectious is very small.  This is also true for the other error sets (see the Supplementary Materials for more details).

@@

One interesting feature we see in the results is that RE and rRE can result in very different estimations and CIs, which may seem unintuitive since rRE is a simple reparametrization of RE.  For example, in Figure \ref{fig:ar-r0}, we see that when the true value of $\rr$ is 1, rRE has a much smaller CI than RE, which may be one reason to prefer rRE to RE.  In this figure, we see when $\gamma << \beta$ (e.g. $\rr=60)$, then all estimates for $\hat{\rr}$ fail.  However, when $\beta >> \gamma$, we the estimates (besides IPR) can accurately obtain the true estimate. This figure also helps to show that RE and rRE are robust to extreme values of $\beta$ and $\gamma$ and hence extreme values of $\rr$.  \com{(For R0=1.5, why do RE, rRE, and LL look like 2 and not 1.5?))}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/err_robust.jpeg}
	\caption{Standardized residuals $\frac{\left ( \hat{\rr} - \rr\right ) }{SE \left [\hat{\rr}\right ]}$ for the estimation methods of RE, rRE, and LL faceted by the four error structures.  The dashed line is the true value of $\rr$.}\label{fig:err-rob}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/model_robust.jpeg}
	\caption{Standardized residuals $\frac{\left ( \hat{\rr} - \rr\right ) }{SE \left [\hat{\rr}\right ]}$ for the estimation methods of RE, rRE, and LL faceted by the four model structures (SIR, Poly1, Poly4, and LSIR).  The dashed line is the true value of $\rr$.}\label{fig:mod-rob}
\end{figure}

@@

      In the previous plots, we assume that we are collecting incidence and recovery date every day.  However, it is also possible to observe weekly, monthly, or even quarterly incidence and recovery data, for example.  In Figure \ref{fig:obs}, we explore estimates of $\rr$ using the data collected at different time periods for our baseline data structure with a full year's worth of data.  We find that we can reliably estimate $\hat{\rr}$ regardless of the collection period.  This is likely due to the previously discussed relationship between $\rr$ and the final size of the epidemic.

@@

\subsection{Variance of $\rr$ estimation methods}\label{sec:sim-var-res}
We examine the variance methods used to estimate CIs for our estimates of $\rr$.  From the above results, we saw that methods such as RE, rRE, LL, and MC result in very small CIs.  On the other hand, methods such as IPR have consistently large CIs, to the point of being uninformative.  The other methods CI width seems to depend heavily on the initial parameters used.  Unsurprisingly, larger $\sigma_X$, $\sigma_Y$ values seem to correspond to larger CI sizes.  On the other hand, large $N$ and $T$ values seem to correspond to smaller CI sizes.   Having a larger percentage of initially infected individuals also seems to be associated with a larger CI.  We also see that having a very large $\rr$ value seems to have less effect on creating a large CI than does having a small value of $\rr$.  Perhaps surprisingly, the width of the CI seems to be independent of the error structure.

@@

	\centering
	\begin{tabular}[t]{l|r|r|r|r}
		\hline
		Model & Mean Estimate & Std. Dev. & Baseline Estimate & Baseline SE\\
		\hline
		RE & 2.0000 & 0.0003 & 1.9999 & 0.0058\\
		\hline
		rRE & 2.0000 & 0.0003 & 1.9997 & 0.0052\\
		\hline
		LMA & 1.8497 & 0.3339 & 1.9974 & 0.3900\\
		\hline
		LMAT & 1.4562 & 0.3577 & 1.3980 & 0.6028 \\
		\hline
		IPR & 4.5214 & 0.3705 & 2.5930 & 9.0408\\
		\hline
		SIPR & 1.7973 & 0.0951 & 0.9251 & $<$ 1e-04 \\
		\hline
		LL & 2.0000 & 0.0003 & 1.9998 & 0.0002\\
		\hline
		MC & 0.9486 & $<$ 1e-04 & 0.9486 & $<$ 1e-04\\
		\hline
		SB & 1.1330 & 0.4163 & 1.0000 & 0.0523\\
		\hline
		LBE & 8.5926 & 6.9421 & 14.6784 & 1.2330\\
		\hline
	\end{tabular}
        \caption{Mean $\rr$ Estimates and Std. Devs, Delta Method/Posterior}
        \label{tab:rep-samp}
\end{table}

@@

	\centering
	\begin{tabular}[t]{l|r|r|r|r}
		\hline
		Model & Mean Estimate & Std. Dev. & Baseline Estimate & Baseline SE\\
		\hline
		RE & 1.9999 & 0.0003 & 1.9999 & 0.0058\\
		\hline
		rRE & 1.9998 & 0.0003 & 1.9997 & 0.0052\\
		\hline
		LMA & -0.9975 & 21.7161 & 1.9974 & 0.3900\\
		\hline
		LMAT & 1.5138 & 1.0485 & 1.3980 & 0.6028 \\
		\hline
		SIPR & 0.7762 & 2.6624 & 0.9252 & $<$ 1e-04 \\
		\hline
		LBE & 19.9935 & 10.640 & 14.6784 & 1.2330 \\
		\hline
	\end{tabular}
	\caption{Mean $\rr$ Estimates and Std. Devs, Block Bootstrap}
	\label{tab:bb-samp}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{images/h1n1-sir.pdf}
  \caption{Data from H1N1 influenza 2009 in the USA.  Weekly wILI is reported, which we transform into S, I, and R counts shown for weeks (start of weeks: April 16-Aug. 27)  using $\gamma^{-1} = 3/7$ weeks.  The observations between the dotted lines denote the points used to estimate $\hat{\rr}$ in \cite{towers2009}.}
  \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{images/h1n1.jpg}
	\caption{Results from SIR models on H1N1 data.}
\end{figure}

@@

The results of this analysis are that most of our methods yield point estimates of $\hat{\rr}$ of about 1.2, except for SB and IPR.  The methods RE, rRE, LMA, LMAT, LL, and MC all have small CIs as well.  These estimates show that one, H1N1 was indeed an epidemic in the time period but are also smaller compared to the Towers and Feng estimate.  As Towers and Feng used a different method of estimating $\rr$ than the above, this shows the importance of testing modelling assumptions when using on epidemic data.

@@

In Table \ref{tab:r0-real-ex}, we provide estimates for $\rr$ that have appeared over the past three decades for HIV, Zika, Ebola, influenza, H1N1 influenza, and measles in different locations throughout the world.  Each of these estimates is greater than $1$, which makes sense as we tend observe prolonged outbreaks of a disease as opposed to isolated occurences.  The largest $\rr$ point estimate we have seen is for measles at $\rr=17$ in England in the 20th century, and the smallest is $\rr=1.08$ for HIV in Portugal in the 21st century.
%% r0 as a property of the model
The quantity $\rr$ may summarize an epidemic, accounting for temperature, seasonality, location, and genetics of a disease, but it is still a property of the model.  Namely, how individuals progress through an epidemic (e.g. susceptible to infectious to recovered compartments) influences how we view $\rr$.  We thus should compare estimates for $\rr$ for a fixed epidemic model.  Here, we limit our analysis to estimates of $\rr$ derived from the SIR model.  Then $\rr = \frac{\beta}{\gamma}$ where $\beta$ is the rate of infection and $\gamma$ is the rate of recovery.  As a consequence, we need to estimate $\rr$ from not only the number of infections but also the number of individuals recovering at each time step.

%% the methods we chose and why we chose them
Additionally, the SIR model assumes monotoncity of S and R compartments which can make estimating $\rr$ more difficult, especially in terms of the CIs since the usual assumption of Gaussian noise may be violated due to the non-symmetric nature of the noise.  We examine \wxxsir methods to estimate $\rr$ which include least squares optimization, polynomial approximations, estimates from incidence and variance, a log-linear model, a Markov chain method, and a simple Bayesian method.  Although, this is not an exhaustive list of methods used to estimate $\rr$ from the SIR model, we believe it covers common and popular methods used throughout the decades.  We overview the details of methods in Section \ref{sec:methods}.

@@

The SIR estimation methods generally give reliable estimates when the data are from an underlying SIR model with little or no noise. However, the estimates from the various methods are generally much worse when the magnitude of the noise is larger, when the parameters of the model are unusual, and when the model is misspecified. We also see a large difference between the $\rr$ estimates and standard errors on the exact same data sets as well, even though all derived from the same model. Thus, when estimating $\rr$ based on an SIR or any other kind of compartment model, we need to take careful consideration of all known factors.

@@

>>>>

\noindent \textbf{Gaussian-Monotone-PAVA (GP) ($\hat{X}_{GP}$, $\hat{Y}_{GP}$, $\hat{Z}_{GP}$)}
\begin{align*}
 \hat{X}_{GP}(t) &= \textnormal{PAVA}(\hat{X}_G(t)) \\
  \hat{Y}_{GP}(t) &= N - \hat{X}_{GP}(t) - \hat{Z}_{GP}(t) \\
  \hat{Z}_{GP}(t) &= \textnormal{PAVA}(\hat{Z}_G(t)) \\
\end{align*}
\textbf{Autoregressive-Monotone-PAVA (ARP) ($\hat{X}_{ARP}$, $\hat{Y}_{ARP}$, $\hat{Z}_{ARP}$)}
\begin{align*}
  \hat{X}_{ARP}(t) &= \textnormal{PAVA}(\hat{X}_{AR}(t)) \\
  \hat{Y}_{ARP}(t) &= N - \hat{X}_{ARP}(t) - \hat{Z}_{ARP}(t) \\
  \hat{Z}_{ARP}(t) &= \textnormal{PAVA}(\hat{Z}_{AR}(t)) 
\end{align*}
This gives us a total of four error structures for each set of conditions  (1) Gaussian (G), (2) Gaussian monotone - PAVA (GP),  (3) autoregressive (AR), and (4) autoregressive monotone - PAVA (ARP) for a given set of parameters $\left (\beta, \gamma, T, X(0), Y(0), Z(0), \sigma_X, \sigma_Y, N \right )$, the infection rate; recovery rate; total number of time points; initial conditions for number of susceptibles, infectious, and recovered individuals; the standard deviations for both the susceptible and infectious compartments; and the total population, respectively.

\subsection{Simulated SIR Data}

We first generate ``baseline'' simulated data from the SIR model under the following conditions shown in Table \ref{tab:baseline-params}:

  % Please add the following required packages to your document preamble:
  % \usepackage{booktabs}
  % \usepackage{graphicx}
  \begin{table}[H]
    \resizebox{\textwidth}{!}{%
      \begin{tabular}{@{}lllllllllll@{}}
        \toprule
        \textbf{Parameter} & $\beta$ & $\gamma$ & $\rr$ & $T$ & $X(0)$ & $Y(0)$ & $Z(0)$ & $\sigma_X$ & $\sigma_Y$ & Error Type \\ \midrule
        \textbf{Value}     & 0.06    & 0.03     & 2.00  & 365 & 99950  & 50     & 0      & 100        & 5          & Gaussian   \\ \bottomrule
      \end{tabular}%
    }
    \caption{Parameter values for baseline simulated data.}
    \label{tab:baseline-params}
  \end{table}
      The baseline data set is generated by adding noise to the susceptible and infectious curves from deterministic SIR paths, which are shown in Figure \ref{fig:baseline-data}.  The recovered compartment is then adjusted to maintain a constant population value.  In this scenario, we see that the infection begins slowly but over the course of a year infects nearly 75\% of the population.
      \begin{figure}
        \centering
        \includegraphics[width=.8\textwidth]{images/baseline_sir.pdf}
        \caption{SIR (XYZ) curves for the deterministic model with $N=100,000$, $(\beta, \gamma) = (0.06, 0.03)$, $(X(0), Y(0), Z(0))= (99950, 50,0)$  The baseline data is generated by using this model and adding noise to the $X$ and $Y$ compartments and adjusting $Z$ at each time step to maintain a constant population $N$.}\label{fig:baseline-data}
        \end{figure}
      The parameters for the baseline simulated data is chosen due to its moderate $\rr$ size of 2, which indicates an outbreak of an epidemic similar to ZIK-V, EVD, and H1N1 pandemic influenza (see Table \ref{tab:r0-real-ex}).  The infection rate and recovery rate ($\beta, \gamma$) are relatively small.  The duration of the disease, under the true model, spans nearly a whole year (see Figure \ref{fig:baseline-data}) and so we use a year's worth of daily observations of ($\hat{X},\hat{Y}, \hat{Z}$) values.  Initially, 50 (0.05\%) infectious individuals are introduced to the population of 100,000 individuals (approximately the average population size of a US county in 2015).  The standard deviation of the $\hat{X}$ values is 0.1\% of the magnitude of the initial number of susceptible individuals, and the standard deviation of the $\hat{Y}$ values is 10\% of the magnitude of the initial number of infectious individuals.  In this setting, the number of infectious individuals increases rapidly.  We use Gaussian error as our baseline error because of its well known statistical properties so we can more easily compare methods to one another.  Later on, we explore the autoregressive and PAVA simulated data, which may better simulate data from  a real epidemic.

@@

After generating our baseline simulated data, we generate different SIR data by changing one of the parameters for simulation each time and using the baseline simulated data as our basis for comparison. The different data sets for a given error structure are shown in Table \ref{tab:simulated-data}.  We repeat  Table \ref{tab:simulated-data}, replacing the ``Error'' feature with  one of the other three error structures for a total of 96 simulated data sets.

@@

\section{Results}\label{sec:results-top}
\subsection{$\rr$ estimation methods}\label{sec:results}
There are numerous aspects to consider when using a method to estimate $\rr$.  Among these include values of both the infection and recovery rate $(\beta, \gamma)$, the number of data points available ($T$), the total size of the population ($N$), the initial percent of susceptible individuals $\left (\frac{X(0)}{N}\right)$, the initial percent of infectious individuals $\left (\frac{Y(0)}{N}\right )$, the magnitude of the variance of both the number of susceptibles and the number of infectious ($\sigma_X^2, \sigma_Y^2$), the error modeling assumptions, the structure of the ``true'' model, and finally methods of estimating the variance.  For the sake of brevity, we do not provide the full results here, and instead examine the results at a high-level.  However, for those interested readers, we provide results in the form of forest plots and tables of estimates of $\rr$ for each data set in the Supplementary Material.  We estimate $\rr$ on the baseline simulated data for each of the \wxxsir methods.  The results are displayed in Figure \ref{fig:baseline-res} and Table \ref{tab:baseline-res}.
\begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/BaseBase.jpg}
  \caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data set $(\beta=.06, \gamma=.03, O=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical, dashed line is the true $\rr$ value.}\label{fig:baseline-res}
  \end{figure}

@@

Overall, in Figure \ref{fig:baseline-res}, we see a wide variation in the estimates of and the standard errors for the estimates of $\rr$ depending on the method used.  The methods of RE, rRE, LMA, LMAT, IPR, and LL all result in a CI that covers the true value of $\rr=2$.  Of these, LMAT and IPR result in a CI which contains 1, so we would not be able to say whether or the outbreak of a disease was due to random chance or not. On the other hand, many methods appear to underestimate the true $\rr$, and many of them have larger standard errors as well.  We would say that IPR's CI is uninformative. Additionally, MC and SIPR have standard errors of less than $.0001$, which is unusually small.  LBE, which assumes that the X and Z compartments are monotonic, overestimates $\rr$

From examining the full results in the Supplementary Material, the results of the baseline simulated data with normal error seem to be representative of the methods as a whole.  In Figure \ref{fig:baseline-res}, we see that the best methods to estimate $\rr$ are RE, rRE, and LL.  These three methods consistently outperform the other methods in the task of accurately estimating $\rr$ with a reasonably small CI when varying a single parameter and holding the others as fixed from the simulated data sets.

In particular, MC and SB almost never capture the true estimate and typically underestimate the true value of $\rr$.  This leads us to believe that these models are for estimating $\rr$ for the SIR model.  Both models rely on parametric assumptions of simple underlying statistical distributions for the course of infection (Binomial and Poisson, respectively) with time-invariant transmission parameters ($\alpha$ and $\lambda$).   That is not to say that variants of these models are unviable for modelling $\rr$, but they do not make for good defaults, as they seem to be too simple to accurately model data from a deterministic SIR model.  It is also an issue that these two methods consistently underestimate $\rr$, which could possibly lead to an unchecked epidemic.

On the other hand, LBE tends to have larger CIs and does not often capture the true value of $\rr$ when the the X and Z compartments are not monotonic functions.  This method tends to overestimate the value of $\rr$.

@@

Of the three best-performing methods, RE and rRE seem to be  more robust to parameter variation than LL.  This is mostly due to the fact that LL produces such small CIs that even a small discrepancy in estimation of $\rr$ can cause the CI to not capture the true value of $\rr$.

However, we do see instances where LL completely outperforms RE and rRE such as in the case where the initial percent of infectious $\frac{I(0)}{N} \cdot 100\%$ is less than $0.01$\%. In Figure \ref{fig:ar-small-I}, we examine the data set generated with autoregressive monotonic error, our most stringent data structure, with varying levels of initial percent of infectious individuals.  When $\frac{I(0)}{N} \cdot 100\%$ is less than $ 0.01$\%, RE and rRE tend to overestimate $\rr$ and have large CIs.  However, LL accurately estimates $\rr$ even when the initial percentage of infectious is very small.  This is also true for the other error sets (see the Supplementary Materials for more details).

@@

One interesting feature we see in the results is that RE and rRE can result in very different estimates and CIs, which may seem unintuitive since rRE is a reparametrization of RE.  For example, in Figure \ref{fig:ar-r0}, we see that when the true value of $\rr$ is 1, rRE has a much smaller CI than RE, which may be one reason to prefer rRE to RE.  In this figure, we see that when $\gamma << \beta$ (e.g. $\rr=60)$, then all estimates for $\hat{\rr}$ fail.  However, when $\beta >> \gamma$, we the estimates (besides IPR) can accurately obtain the true estimate. This figure also helps to show that RE and rRE are robust to extreme values of $\beta$ and $\gamma$ and hence extreme values of $\rr$.  \com{(For R0=1.5, why do RE, rRE, and LL look like 2 and not 1.5?))}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/err_robust.jpeg}
	\caption{Standardized residuals $\frac{\left ( \hat{\rr} - \rr\right ) }{SE \left [\hat{\rr}\right ]}$ for the estimation methods of RE, rRE, and LL grouped by the four error structures.  The dashed line is the true value of $\rr$. \com{X axis should be standardized R0 residuals.  Same for Figure 9.  Try increasing element\_text size}}\label{fig:err-rob}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/model_robust.jpeg}
	\caption{Standardized residuals $\frac{\left ( \hat{\rr} - \rr\right ) }{SE \left [\hat{\rr}\right ]}$ for the estimation methods of RE, rRE, and LL grouped by the four model structures (SIR, Poly1, Poly4, and LSIR).  The dashed line is the true value of $\rr$.}\label{fig:mod-rob}
\end{figure}

@@

      In the previous plots, we assume that we are collecting incidence and recovery data every day.  However, it is also possible to observe weekly, monthly, or even quarterly incidence and recovery data, for example.  In Figure \ref{fig:obs}, we explore estimates of $\rr$ using the data collected at different time periods for our baseline data structure with a full year's worth of data.  We find that we can reliably estimate $\hat{\rr}$ regardless of the collection period.  This is likely due to the previously discussed relationship between $\rr$ and the final size of the epidemic.

@@

\subsection{Variance of $\rr$ estimation methods}\label{sec:sim-var-res}
We examine the variance methods used to estimate CIs for our estimates of $\rr$.  From the above results, we saw that methods such as RE, rRE, LL, and MC result in very small CIs.  On the other hand, methods such as IPR have consistently large CIs, to the point of being uninformative.  For the other methods, the CI width seems to depend heavily on the parameters used to simulate the data.  Unsurprisingly, larger $\sigma_X$, $\sigma_Y$ values seem to correspond to larger CI sizes.  On the other hand, large $N$ and $T$ values seem to correspond to smaller CI sizes.   Having a larger percentage of initially infected individuals also seems to be associated with a larger CI.  We also see that having a very large $\rr$ value seems to have less effect on creating a large CI than does having a small value of $\rr$.  Perhaps surprisingly, the width of the CI seems to be independent of the error structure.

@@

	\centering
	\begin{tabular}[t]{l|r|r|r|r}
		\hline
		Model & Empirical Mean Est. & Empirical Std. Dev. & Method Mean & Method SE\\
		\hline
		RE & 2.0000 & 0.0003 & 1.9999 & 0.0058\\
		\hline
		rRE & 2.0000 & 0.0003 & 1.9997 & 0.0052\\
		\hline
		LMA & 1.8497 & 0.3339 & 1.9974 & 0.3900\\
		\hline
		LMAT & 1.4562 & 0.3577 & 1.3980 & 0.6028 \\
		\hline
		IPR & 4.5214 & 0.3705 & 2.5930 & 9.0408\\
		\hline
		SIPR & 1.7973 & 0.0951 & 0.9251 & $<$ 1e-04 \\
		\hline
		LL & 2.0000 & 0.0003 & 1.9998 & 0.0002\\
		\hline
		MC & 0.9486 & $<$ 1e-04 & 0.9486 & $<$ 1e-04\\
		\hline
		SB & 1.1330 & 0.4163 & 1.0000 & 0.0523\\
		\hline
		LBE & 8.5926 & 6.9421 & 14.6784 & 1.2330\\
		\hline
	\end{tabular}
        \caption{Mean $\rr$ Estimates and Std. Devs, Delta Method/Posterior}
        \label{tab:rep-samp}
\end{table}

@@

	\centering
	\begin{tabular}[t]{l|r|r|r|r}
          \hline
          	Model & Empirical Mean Est. & Empirical Std. Dev. & Method Mean & Method SE\\
		\hline
		RE & 1.9999 & 0.0003 & 1.9999 & 0.0058\\
		\hline
		rRE & 1.9998 & 0.0003 & 1.9997 & 0.0052\\
		\hline
		LMA & -0.9975 & 21.7161 & 1.9974 & 0.3900\\
		\hline
		LMAT & 1.5138 & 1.0485 & 1.3980 & 0.6028 \\
		\hline
		SIPR & 0.7762 & 2.6624 & 0.9252 & $<$ 1e-04 \\
		\hline
		LBE & 19.9935 & 10.640 & 14.6784 & 1.2330 \\
		\hline
	\end{tabular}
	\caption{Mean $\rr$ Estimates and Std. Devs, Block Bootstrap.  \com{Shouldn't we get predictions from LL?}}
	\label{tab:bb-samp}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{images/h1n1-sir.pdf}
  \caption{\com{Change title to ``Observed SIR...''}Observed data from H1N1 influenza 2009 in the USA.  Weekly wILI is reported, which we transform into S, I, and R counts shown for weeks (start of weeks: April 16-Aug. 27)  using $\gamma^{-1} = 3/7$ weeks.  The observations between the dotted lines denote the points used to estimate $\hat{\rr}$ in \cite{towers2009}.}
  \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{images/h1n1.jpg}
	\caption{\com{Just show 0-12 R0.  Need LBE in table}Results from SIR models on H1N1 data.}
\end{figure}

@@

The results of this analysis are that most of our methods yield point estimates of $\hat{\rr}$ of about 1.2, except for SB and IPR.  The methods RE, rRE, LMA, LMAT, LL, and MC all have small CIs as well.  Using RE, we are 95\% confident that $\hat{\rr} \in [0.48, 1.89]$, which is in agreement with the Towers and Feng estimate.  However, when using rRE we are 95\% confident that $\hat{\rr} \in [1.14, 1.23]$ and for LL $\hat{\rr} \in [1.15, 1.22]$, which is much smaller than their estimate of $\rr$.  These estimates show that one, H1N1 was indeed an epidemic in the time period but are also smaller compared to the Towers and Feng estimate.  As Towers and Feng used a different method of estimating $\rr$ than the above, this shows the importance of testing modelling assumptions when using on epidemic data.

@@

In Table \ref{tab:r0-real-ex}, we provide estimates for $\rr$ that have appeared in the literature over the past three decades for HIV, Zika, Ebola, influenza, H1N1 influenza, and measles in different locations throughout the world.  Each of these estimates is greater than $1$, which makes sense as we tend to observe prolonged outbreaks of a disease as opposed to isolated occurences.  The largest $\rr$ point estimate we have seen is for measles at $\rr=17$ in England in the 20th century, and the smallest is $\rr=1.08$ for HIV in Portugal in the 21st century.
%% r0 as a property of the model
The quantity $\rr$ may summarize an epidemic, accounting for temperature, seasonality, location, and genetics of a disease, but it is still a property of the model.  Namely, how individuals progress through an epidemic (e.g. susceptible to infectious to recovered compartments) influences how we view $\rr$.  We thus should compare estimates for $\rr$ for a fixed epidemic model.  Here, we limit our analysis to estimates of $\rr$ derived from the SIR model.  Then $\rr = \frac{\beta}{\gamma}$ where $\beta$ is the rate of infection and $\gamma$ is the rate of recovery.  As a consequence, we need to estimate $\rr$ from not only the number of infections but also the number of individuals recovering at each time step.

%% the methods we chose and why we chose them
Additionally, the SIR model assumes monotoncity of S and R compartments which can make estimating $\rr$ more difficult, especially in terms of the CIs since the usual assumption of Gaussian noise is violated due to the non-symmetric nature of the noise.  We examine \wxxsir methods to estimate $\rr$ which include least squares optimization, polynomial approximations, estimates from incidence and variance, a log-linear model, a Markov chain method, and a simple Bayesian method.  Although, this is not an exhaustive list of methods used to estimate $\rr$ from the SIR model, we believe it covers common and popular methods used throughout the decades.  We reviewed the details of methods in Section \ref{sec:methods}.

@@

The $\rr$ estimation methods generally give reliable estimates when the data are from an underlying SIR model with little or no noise. However, the estimates from the various methods are generally much worse when the magnitude of the noise is larger, when the parameters of the model are unusual, and when the model is misspecified.  Thus, when estimating $\rr$ based on an SIR or any other kind of compartment model, we need to take careful consideration of all known factors and carefully state and verify the assumptions of the chosen model(s).

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
